Temporal trends (NAMCS)

### 2017-07-24 ms1/namcs_trends.Rmd
re-streamlined

```{r load_data}
# Caveat: I removed a stratum from the 2011 data that had only one PSU in it.
# Also: I'm not sure that combining years like this is going to be kosher.

years = 2011:2014
n_years = length(years)
mpy = function(x) x/(1e6 * n_years) # millions per year

@@

census = read_tsv('../../../../db/census/census_totals.tsv')

@@

visits = read_tsv_years('../../db/namcs/data/namcs_visits_%i.tsv') %>%
  mutate(age_group=case_when(.$age <= 19 ~ 'child',
                             .$age <= 64 ~ 'adult',
                             .$age <= 92 ~ 'elderly')) %>%
  left_join(dx_visit, by='visit_id') %>%
  replace_na(list(acute_rc=FALSE)) %>%
  left_join(n_abx, by='visit_id') %>% replace_na(list(n_abx=0)) %>%
  left_join(abx_visit, by='visit_id') %>%
  mutate_at(top_abx$abx, function(x) if_else(is.na(x), 0L, x))

@@

I used the 2011-2014 NAMCS data. All numbers of prescriptions will be presented in millions per year. I'll mostly consider only the adults.

@@

## Summary of trends

### Overall trends

It looks by eye like the total number of antibiotic prescriptions per capita is going down:

```{r overall}
overall_trend = svyby(~n_abx, ~year, design=design, svytotal) %>%
  as_data_frame() %>%
  left_join(census, by='year') %>%
  mutate(abx_per_capita=n_abx/n_total, se_per_capita=se/n_total)

overall_trend %>%
  mutate(ci=1.96*se_per_capita, ymin=abx_per_capita-ci, ymax=abx_per_capita+ci) %>%
  ggplot(aes(x=year, y=abx_per_capita, ymin=ymin, ymax=ymax)) +
  geom_point() + geom_line() + geom_linerange() +
  theme_minimal()
```

There is no support for a trend per visit:

```{r lm_functions}
# survey poisson model
svy_pm = function(design, frmla) svyglm(frmla, design=design, family='poisson') %>% tidy

# log glm
log_glm = function(df, frmla, weights=NULL) glm(formula=as.formula(frmla), data=df, family=gaussian(link='log'), weights=weights) %>% tidy
```

```{r overall_per_visit}
svy_pm(design, n_abx ~ year) %>% kable
```

Nor for a trend per capita:

```{r overall_per_capita}
overall_trend %>%
  mutate(year=year-min(year)) %>%
  log_glm('abx_per_capita ~ year', weights=1.0/.$se_per_capita^2) %>%
  kable
```

### Trends by drug

#### Per capita

```{r drugs_per_capita_model}
n_abx_f = function(design, a) {
  svyby(as.formula(paste0('~', a)), ~year, design=design, svytotal) %>%
    as_data_frame() %>%
    rename_(.dots=setNames(a, 'n_abx')) %>%
    mutate(abx=a)
}

# apc = antibiotic per capita
apc_f = function(design, denom) {
  lapply(top_abx$abx, function(x) n_abx_f(design, x)) %>% bind_rows() %>%
    left_join(census, by='year') %>%
    rename_(.dots=setNames(denom, 'denom')) %>%
    mutate(apc=n_abx/denom, se=se/denom)
}

abx_time = apc_f(design, 'n_total')
abx_time_adult = apc_f(adult_design, 'n_adult')
```

```{r drugs_per_capita_time}
abx_time %>%
  mutate(ci=1.96*se, ymin=apc-ci, ymax=apc+ci) %>%
  ggplot(aes(x=year, y=apc, ymin=ymin, ymax=ymax)) +
  facet_wrap(~abx, ncol=3, scales='free_y') +
  geom_point() + geom_line() + geom_linerange() +
  theme_minimal()
```

And the same plot again, but with fixed scales:

```{r drugs_time_plot_fixed}
abx_time %>%
  mutate(ci=1.96*se, ymin=apc-ci, ymax=apc+ci) %>%
  ggplot(aes(x=year, y=apc, ymin=ymin, ymax=ymax)) +
  facet_wrap(~abx, ncol=3) +
  geom_point() + geom_line() + geom_linerange() +
  theme_minimal()
```

It doesn't look by eye that any trends are particularly strong. I used weighted
log regression to look for trends in totals:

```{r drugs_per_capita_time_trends}
abx_total_models = abx_time %>%
  group_by(abx) %>%
  do(log_glm(., 'apc ~ year', weights=1.0/.$se^2)) %>%
  ungroup()

abx_total_models %>%
  filter(term=='year') %>% select(-term) %>%
  arrange(estimate) %>%
  kable
```

```{r drugs_per_capita_figure}
bind_rows(
  read_tsv('tables/model_abx.tsv') %>%
    filter(term=='year', model=='poisson') %>%
    left_join(top_abx, by=c('antibiotic'='medicare_abx')) %>%
    mutate(abx=if_else(is.na(abx), antibiotic, abx)) %>%
    select(abx, estimate, std.error) %>%
    mutate(data_source='medicare'),
  abx_total_models %>%
    filter(term=='year') %>%
    select(abx, estimate, std.error) %>%
    mutate(data_source='namcs')
) %>%
  #mutate(abx=factor(abx, levels=top_abx$abx)) %>%
#abx_total_models %>%
#  filter(term=='year') %>%
  mutate(ci=1.96*std.error, y=exp(estimate)-1, ymin=exp(estimate-ci)-1, ymax=exp(estimate+ci)-1) %>%
  #select(abx, y, ymin, ymax) %>%
  mutate_if(is.numeric, function(x) x*100) %>%
  #arrange(y) %>% mutate(abx=forcats::fct_inorder(abx)) %>%
  #ggplot(aes(x=abx, y=y, ymin=ymin, ymax=ymax)) +
  ggplot(aes(x=abx, y=y, ymin=ymin, ymax=ymax, color=data_source)) +
  geom_linerange() +
  geom_point() +
  coord_flip() +
  ylab('pct change in consumption per capita') +
  theme_minimal()
```

#### Per visit

I used Poisson regressions to see if anything was changing *per visit* with
time: levo and amox/clav. (These results are only with adults.)

```{r drugs_per_visit}
f = function(design) {
  lapply(top_abx$abx, function(a) {
    svy_pm(design, as.formula(paste0(a, '~year'))) %>%
      mutate(abx=a)
  }) %>% bind_rows()
}

abx_time_model = f(design)
abx_time_model_adult = f(adult_design)
```

```{r drugs_per_visit_table}
abx_time_model_adult %>%
  filter(term=='year') %>%
  select(abx, estimate, p.value) %>%
  mutate(adj.p.value=p.adjust(p.value, method='BH')) %>%
  mutate(pct=100*(exp(estimate)-1)) %>%
  arrange(desc(pct)) %>%
  kable
```

```{r drugs_per_visit_figure}
abx_time_model_adult %>%
  filter(term=='year') %>%
  mutate(ci=1.96*std.error, y=exp(estimate)-1, ymin=exp(estimate-ci)-1, ymax=exp(estimate+ci)-1) %>%
  select(abx, y, ymin, ymax) %>%
  mutate_if(is.numeric, function(x) x*100) %>%
  arrange(y) %>% mutate(abx=forcats::fct_inorder(abx)) %>%
  ggplot(aes(x=abx, y=y, ymin=ymin, ymax=ymax)) +
  geom_point() + geom_linerange() +
  coord_flip() +
  ylab('pct change in consumption per visit') +
  theme_minimal()
```

```{r special_times}
levo_t1 = abx_time_adult %>% filter(abx=='levo', year==2011)
levo_t2 = abx_time_adult %>% filter(abx=='levo', year==2014)
amoxclav_t1 = abx_time_adult %>% filter(abx=='amoxclav', year==2011)
amoxclav_t2 = abx_time_adult %>% filter(abx=='amoxclav', year==2014)

@@

#### Per visit, Among everyone

I do the same analysis, but among everyone.

```{r drugs_per_visit_everyone}
abx_time_model %>%
  filter(term=='year') %>%
  select(-term) %>%
  kable
```

>>>>

```{r load_namcs_data}
# Caveat: I removed a stratum from the 2011 data that had only one PSU in it.
# Also: I'm not sure that combining years like this is going to be kosher.

years = 2011:2014
n_years = length(years)

@@

@@

visits = read_tsv_years('../../db/namcs/data/namcs_visits_%i.tsv') %>%
  mutate(age_group=case_when(.$age <= 19 ~ 'child',
                             .$age <= 64 ~ 'adult',
                             .$age <= 92 ~ 'elderly')) %>%
  left_join(dx_visit, by='visit_id') %>%
  replace_na(list(acute_rc=FALSE)) %>%
  left_join(n_abx, by='visit_id') %>% replace_na(list(n_abx=0)) %>%
  left_join(abx_visit, by='visit_id') %>%
  mutate_at(top_abx$abx, function(x) if_else(is.na(x), 0L, x))
```

```{r load_other_data}
census = read_tsv('../../../../db/census/census_totals.tsv') %>%
  mutate(n_nonchild=n_total - n_children)

mpy = function(x) x/(1e6 * n_years) # millions per year

@@

I used the 2011-2014 NAMCS data. All numbers of prescriptions will be presented
in millions per year. I'll mostly consider only the adults (i.e., those over 18 years).

@@

This means that visits per American went down. Thus, consumption of some drug
could be increasing per visit but not be increasing per capita, for example.

## Trends per visit

Methodologically, trends per visit is easier, because we just perform
regressions on the records in NAMCS. The con is that the results are more
difficult to interpret in light of the fact that there are trends in population
size as well as in healthcare usage.

In all the results that follow, I'm looking at consumption among the adult
population only.

```{r lm_functions}
add_pct_change = function(df) {
  mutate(df, ci=1.96*std.error,
             pct_change=exp(estimate)-1,
             ci_low=exp(estimate-ci)-1,
             ci_high=exp(estimate+ci)-1) %>%
    mutate_at(vars(pct_change, ci_low, ci_high), function(x) x*100)
}
```

```{r per_visit_model}
# survey poisson model
svy_pm = function(design, frmla) svyglm(frmla, design=design, family='poisson') %>% tidy

overall_pv = adult_design %>% svy_pm(n_abx ~ year) %>% mutate(abx='overall')

drug_pv = top_abx$abx %>%
  lapply(function(a) svy_pm(adult_design, as.formula(paste0(a, '~year'))) %>% mutate(abx=a)) %>%
  bind_rows()
```

```{r per_visit_display}
x = bind_rows(overall_pv, drug_pv) %>%
  filter(term=='year') %>%
  mutate(adj.p.value=p.adjust(p.value, method='BH')) %>%
  add_pct_change() %>%
  arrange(pct_change) %>%
  mutate(abx=forcats::fct_inorder(abx))

ggplot(x, aes(x=abx, y=pct_change, ymin=ci_low, ymax=ci_high)) +
  geom_linerange() +
  geom_point() +
  coord_flip() +
  xlab('') +
  ylab('pct change in consumption per visit') +
  ggtitle('Per visit') +
  theme_minimal()

x %>%
  select(abx, ci_low, pct_change, ci_high, p.value, adj.p.value) %>%
  mutate(sig=if_else(adj.p.value < 0.05, '*', '')) %>%
  kable
```

## Trends per capita

```{r per_capita_counts}
# apc = antibiotics per capita

overall_pc_f = function(design, denom) {
  svyby(~n_abx, ~year, design=design, svytotal) %>%
    as_data_frame() %>%
    left_join(census, by='year') %>%
    rename_(.dots=setNames(denom, 'denom')) %>%
    mutate(apc=n_abx/denom, se=se/denom) %>%
    mutate(abx='overall')
}

n_abx_f = function(design, a) {
  svyby(as.formula(paste0('~', a)), ~year, design=design, svytotal) %>%
    as_data_frame() %>%
    rename_(.dots=setNames(a, 'n_abx')) %>%
    mutate(abx=a)
}

apc_f = function(design, denom) {
  lapply(top_abx$abx, function(x) n_abx_f(design, x)) %>%
    bind_rows() %>%
    left_join(census, by='year') %>%
    rename_(.dots=setNames(denom, 'denom')) %>%
    mutate(apc=n_abx/denom, se=se/denom) %>%
    bind_rows(overall_pc_f(design, denom))
}

per_capita_counts = apc_f(adult_design, 'n_nonchild')
per_capita_counts_all = apc_f(design, 'n_total')
```

```{r per_capita_counts_display}
# free scales, with overall
per_capita_counts %>%
  mutate(abx=factor(abx, levels=c('overall', top_abx$abx))) %>%
  mutate(ci=1.96*se, ymin=apc-ci, ymax=apc+ci) %>%
  ggplot(aes(x=year, y=apc, ymin=ymin, ymax=ymax)) +
  facet_wrap(~abx, ncol=3, scales='free_y') +
  geom_point() + geom_line() + geom_linerange() +
  theme_minimal()

# fixed scales, without overall
per_capita_counts %>%
  filter(abx != 'overall') %>%
  mutate(abx=factor(abx, levels=top_abx$abx)) %>%
  mutate(ci=1.96*se, ymin=apc-ci, ymax=apc+ci) %>%
  ggplot(aes(x=year, y=apc, ymin=ymin, ymax=ymax)) +
  facet_wrap(~abx, ncol=3, scales='fixed') +
  geom_point() + geom_line() + geom_linerange() +
  theme_minimal()
```

```{r per_capita_models}
per_capita = per_capita_counts %>%
  group_by(abx) %>%
  do(tidy(glm(apc ~ year, data=., family=gaussian(link='log'), weights=1.0/se^2)))
```

```{r per_capita_models_display}
x = per_capita %>%
  filter(term=='year') %>%
  mutate(adj.p.value=p.adjust(p.value, method='BH')) %>%
  add_pct_change()

read_tsv('tables/model_abx.tsv') %>%
  filter(term=='year', model=='poisson') %>%
  left_join(top_abx, by=c('antibiotic'='medicare_abx')) %>%
  mutate(abx=if_else(is.na(abx), antibiotic, abx)) %>%
  mutate(data_source='medicare') %>%
  bind_rows(per_capita %>% filter(term=='year') %>% mutate(data_source='namcs')) %>%
  add_pct_change() %>%
  mutate(abx=forcats::fct_reorder(abx, pct_change)) %>%
  ggplot(aes(x=abx, y=pct_change, ymin=ci_low, ymax=ci_high, color=data_source)) +
  geom_linerange() +
  geom_point() +
  coord_flip() +
  xlab('') +
  ylab('pct change in consumption per capita') +
  ggtitle('Per capita') +
  theme_minimal()

x %>%
  select(abx, ci_low, pct_change, ci_high, p.value, adj.p.value) %>%
  arrange(desc(pct_change)) %>%
  kable
```

## Special drugs

```{r special_times}
levo_t1 = per_capita_counts %>% filter(abx=='levo', year==2011)
levo_t2 = per_capita_counts %>% filter(abx=='levo', year==2014)
amoxclav_t1 = per_capita_counts %>% filter(abx=='amoxclav', year==2011)
amoxclav_t2 = per_capita_counts %>% filter(abx=='amoxclav', year==2014)

@@
